Estimating length of SARS-CoV-2 from doubly interval censored data with missed infections

### 2024-03-19 paper-surv-analysis.tex
spelling / grammar check

Let $i \in \set{D}$ indicate $\sched_i$ is associated with a detected episode.
That is, $i \in \set{D}$ if and only if there exists a $j(i)$ such that $b_{j(i)}$ and $e_{j(i)}$ such that $(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode.
$(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode when the following three conditions are met.
First, $\exists t \in \sched_i$ such that $b_{j(i)} \leq t \leq e_{j(i)}$; this condition is equivalent to having at least one positive test for the episode.
Second, $\exists t \in \sched_i$ such that $t < b_{j(i)}$, to lower bound the start of the episode.
Finally, $\exists t \in \sched_i$ such that $t > e_{j(i)}$, to upper bound the end of the episode; this condition will be met for any episode sufficiently far in the past in individuals not lost to follow-up (almost always the case in this context).
For a new context where recent infections need to be considered, this condition can be relaxed by considering these episodes as right-censored, at little cost except a more complicated likelihood expression.

@@

I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
If it is a false negative, then I will consider the episode's end right-censored.
However, if it is a true negative, then the episode's end is interval-censored, as previously.
A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.

@@

First, if $e_{j(i)} \leq r_{j(i)}^{(e)}$, meaning that the test at $r_{j(i)}^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
% In this case, the test at $r_{j(i)}^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
The true negative occurs with probability 1, by the assumption of no false positives.
\begin{align}
&p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \theta) \\
&= p(\vec{y_{j(i)}}, l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\ % &\text{as no false positives}
&= p(\vec{y_{j(i)}} \mid l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \theta) p(l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right)
\label{imperf-test:eq:ll-ei-lt-ri}
\end{align}

@@

To generate the durations, an assumption is required for the distribution of the duration of positivity.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
The tail is modified based on a previously unpublished analysis of CIS data, without accounting for several of the issues previously described.
Walker uses survival analysis to estimate the duration from CIS data.
The initiating event is assumed known as the time the episode was detected.
The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.

>>>>

Let $i \in \set{D}$ indicate $\sched_i$ is associated with a detected episode.
That is, $i \in \set{D}$ if and only if there exists a $j(i)$ such that $b_{j(i)}$ and $e_{j(i)}$ such that $(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode.
$(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode when the following three conditions are met.
First, $\exists t \in \sched_i$ such that $b_{j(i)} \leq t \leq e_{j(i)}$; this condition is equivalent to having at least one positive test for the episode.
Second, $\exists t \in \sched_i$ such that $t < b_{j(i)}$, to lower bound the start of the episode.
Finally, $\exists t \in \sched_i$ such that $t > e_{j(i)}$, to upper bound the end of the episode; this condition will be met for any episode sufficiently far in the past in individuals not lost to follow-up (almost always the case in this context).
For a new context where recent infections need to be considered, this condition can be relaxed by considering these episodes as rightcensored, at little cost except a more complicated likelihood expression.

@@

I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
If it is a false negative, then I will consider the episode's end right-censored.
However, if it is a true negative, then the episode's end is interval censored, as previously.
A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.

@@

First, if $e_{j(i)} \leq r_{j(i)}^{(e)}$, meaning that the test at $r_{j(i)}^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
% In this case, the test at $r_{j(i)}^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
The true negative occurs with probability 1, by the assumption of no false positives.
\begin{align}
&p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \theta) \\
&= p(\vec{y_{j(i)}}, l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\ % &\text{as no false positives}
&= p(\vec{y_{j(i)}} \mid l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \theta) p(l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right)
\label{imperf-test:eq:ll-ei-lt-ri}
\end{align}

@@

To generate the durations, an assumption is required for the distribution of the duration of positivity.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
The tail is modified based on a previously unpublished analysis of CIS data, without accounting for several of the issues previously described.
Walker uses survival analysis to estimate the duration from CIS data.
The initiating event is assumed known as the time the episode was detected.
The final event is assumed interval censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.
